Data Analysis

### 2025-08-07 docs/code/07_data_analysis.Rmd
First mansucript draft

@@

@@

6\. Blinding: (Levels: yes, no(

7\. Random Assignment: (Levels: yes, no)

8\. Missing Data: (Levels: yes, no)

Note that the specific subset of data for each meta-regression may differ based on the data availability (e.g., information about the type of parasites is only available for effect sizes based on parasite/pathogen related fitness proxy). We will conduct meta-regression on suitable subsets of data for each analysis, provided we have sufficient effect sizes per moderator level (i.e., at least 5 effect sizes per moderator level).

::: callout-important
## Phylogenetic Effects

Note also that if we obtain additional effect sizes for other bird species after we update our search and/or contact authors, our models may need to account for phylogenetic nonindependence via the inclusion of two additional random effects: "species" and "phylogeny" following Cinar et al. (2022).
:::

## Other Analysis

### Risk of Bias Assessment

-   RoB - The outcome will be presented in the form of a [**summary table**]{.underline}. We will conduct uni-moderator meta-regressions using these variables.

@@

## --- Now for SMDH -----
dataset_SMDH_published<-dataset_SMDH_published%>%
  mutate(year_MeanCenter=scale(as.numeric(year_publication),scale=F)[,1])

@@

```{r}
# ## For lnRR complete dataset
# 
# ## I am changing the year of unpublished studies to 2025 which is to now, assuming that if they were to be published now, what would be the effect. I will keep in mind that in this model if I do find an evidence of publication bias over years, I should keep in mind it is non-independent of the unpublished studies
# 
# dataset_lnRR_all_in<-dataset_lnRR%>%mutate(year_publication = case_when(is.na(year_publication) ~ 2025, TRUE ~ year_publication))%>%
#   mutate(year_MeanCenter=scale(as.numeric(year_publication),scale=F)[,1])%>%
#   mutate(effective_n_inv=1/((4*effective_n_experiment*effective_n_control)/(effective_n_experiment+effective_n_control)))%>%
#   mutate(sqrt_effective_n_inv=sqrt(effective_n_inv))%>%
#   mutate(comparision_type= case_when(is.na(comparision_type) ~ 4, TRUE ~ comparision_type))
# 
# VCV_lnRR_all_in<- vcalc(vi = lnRR_variance,
#                   cluster = paper_ID,
#                   obs = Observation_ID,
#                   data = dataset_lnRR_all_in,
#                   rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
#                   )
# 
# lnRR_all_in <- rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
#                     V = VCV_lnRR_all_in, # specify lnRR's sampling variance-covariance matrix;
#                     mods = ~ time_of_gnm_addition + 
#                       year_MeanCenter + 
#                       sqrt_effective_n_inv + 
#                       as.factor(comparision_type) + 
#                       bird_species + 
#                       trait_type,
#                     random = list(~ 1 | paper_ID,
#                                   ~ 1 | Observation_ID,
#                                   ~ 1 | group_ID_coded,
#                                   ~ 1 | population_ID,
#                                   ~ 1 | bird_species),
#                     method = "REML",
#                     control = list(optimizer="optim"),
#                     test = "t",
#                     data = dataset_lnRR_all_in)
# 
# 
# 
# ## For SMDH complete published dataset
# 
# dataset_SMDH_all_in<-dataset_SMDH%>%mutate(year_publication = case_when(is.na(year_publication) ~ 2025, TRUE ~ year_publication))%>%
#   mutate(year_MeanCenter=scale(as.numeric(year_publication),scale=F)[,1])%>%
#   mutate(effective_n_inv=1/((4*effective_n_experiment*effective_n_control)/(effective_n_experiment+effective_n_control)))%>%
#   mutate(sqrt_effective_n_inv=sqrt(effective_n_inv))%>%
#   mutate(comparision_type= case_when(is.na(comparision_type) ~ 4, TRUE ~ comparision_type))
# 
# VCV_SMDH_all_in<- vcalc(vi = SMDH_variance,
#                   cluster = paper_ID,
#                   obs = Observation_ID,
#                   data = dataset_SMDH_all_in,
#                   rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
#                   )
# 
# SMDH_all_in<- rma.mv(yi = SMDH_sign, # specify lnRR as the effect size measure;
#                     V = VCV_SMDH_all_in, # specify lnRR's sampling variance-covariance matrix;
#                     mods = ~ time_of_gnm_addition + 
#                       year_MeanCenter + 
#                       sqrt_effective_n_inv + 
#                       as.factor(comparision_type) + 
#                       bird_species + 
#                       trait_type,
#                     random = list(~ 1 | paper_ID,
#                                   ~ 1 | Observation_ID,
#                                   ~ 1 | group_ID_coded,
#                                   ~ 1 | population_ID,
#                                   ~ 1 | bird_species),
#                     method = "REML",
#                     control = list(optimizer="optim"),
#                     test = "t",
#                     data = dataset_SMDH_all_in)
```

@@

including including post-hoc Wald tests

For this I have to run all the meta-regressions again without removing the intercepts, good luck to me!

>>>>

Are the meta-regression robust to this analysis

dataset_lnRR_design_flagged_no_missingES <- lnRR_flagged_no_missingES %>% filter(!is.na(comparision_type))

VCV_lnRR_design_flagged_no_missingES <- vcalc(vi = lnRR_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_lnRR_design_flagged_no_missingES,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )
lnRR_design_flagged_no_missingE <-
  metafor::rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
                    V = VCV_lnRR_design_flagged_no_missingES ,# specify lnRR's sampling variance;
                    mods = ~ - 1 + comparision_type,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                  test = "t",
                  control = list(optimizer="optim"),
                  data = dataset_lnRR_design_flagged_no_missingES)

# This analysis only includes dataset that has a proxy relating to comaparision type
dataset_SMDH_design_flagged_no_missingES <-SMDH_flagged_no_missingES %>% filter(!is.na(comparision_type))

VCV_SMDH_design_flagged_no_missingES <- vcalc(vi = SMDH_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_SMDH_design_flagged_no_missingES,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )
SMDH_design_flagged_no_missingE <-
  metafor::rma.mv(yi = SMDH_sign, # specify lnRR as the effect size measure;
                    V = VCV_SMDH_design_flagged_no_missingES ,# specify lnRR's sampling variance;
                    mods = ~ - 1 + comparision_type,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                  test = "t",
                  control = list(optimizer="optim"),
                  data = dataset_SMDH_design_flagged_no_missingES)

@@

# Sensitivity Analysis: Excluding Non-aromatic Vs. Aromatic

We are aware that removing over 50% of the dataset will cause some moderator levels to drop out. But if our goal is to critically examine the role of aromaticity and potentially argue that its importance has been overstated, this could be a very informative step.

## For the lnRR effect size...
dataset_lnRR_no_comp1<-dataset_lnRR%>%filter(comparision_type!=1 | is.na(comparision_type))
## Only some rows are removed... 
nrow(dataset_lnRR)-nrow(dataset_lnRR_no_comp1)

VCV_lnRR_no_comp1 <- vcalc(vi = lnRR_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_lnRR_no_comp1,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )

SA_lnRR_no_comp1<- rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
                    V = VCV_lnRR_no_comp1, # specify lnRR's sampling variance-covariance matrix;
                    mods = ~ 1,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                    control = list(optimizer="optim"),
                    test = "t",
                    data = dataset_lnRR_no_comp1)

# Saving the model
saveRDS(SA_lnRR_no_comp1, file = here::here("model/sensitivity_analysis/SA_lnRR_no_comp1.rds"))

# saving the dataset for this model
write.csv(dataset_lnRR_no_comp1,here::here("data/05_data_sensitivity_analysis/dataset_lnRR_no_comp1.csv"))

dataset_SMDH_no_comp1<-dataset_SMDH%>%filter(comparision_type!=1 | is.na(comparision_type))
## Only some rows are removed... 
nrow(dataset_SMDH)-nrow(dataset_SMDH_no_comp1)

VCV_SMDH_no_comp1 <- vcalc(vi = SMDH_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_SMDH_no_comp1,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )

SA_SMDH_no_comp1<- rma.mv(yi = SMDH_sign, # specify lnRR as the effect size measure;
                    V = VCV_SMDH_no_comp1, # specify lnRR's sampling variance-covariance matrix;
                    mods = ~ 1,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                    control = list(optimizer="optim"),
                    test = "t",
                    data = dataset_SMDH_no_comp1)

# Saving the model
saveRDS(SA_SMDH_no_comp1, file = here::here("model/sensitivity_analysis/SA_SMDH_no_comp1.rds"))

# saving the dataset for this model
write.csv(dataset_SMDH_no_comp1,here::here("data/05_data_sensitivity_analysis/dataset_SMDH_no_comp1.csv"))

#----------------------------------------------------------------------------
# Rerunning the two mechanistic hypothesis meta-regressions with those effect sizes excluded
#----------------------------------------------------------------------------

lnRR_hypothesis_no_comp1 <-
  metafor::rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
                    V = VCV_lnRR_no_comp1,# specify lnRR's sampling variance;
                    mods = ~ - 1 + Hypothesis,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                  test = "t",
                  control = list(optimizer="optim"),
                  data = dataset_lnRR_no_comp1)
# Save model
saveRDS(lnRR_hypothesis_no_comp1, file = here::here("model/lnRR_hypothesis_no_comp1.rds"))

SMDH_hypothesis_no_comp1 <-
  metafor::rma.mv(yi = SMDH_sign, # specify lnRR as the effect size measure;
                    V = VCV_SMDH_no_comp1,# specify lnRR's sampling variance;
                    mods = ~ - 1 + Hypothesis,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                  test = "t",
                  control = list(optimizer="optim"),
                  data = dataset_SMDH_no_comp1)

# Save model
saveRDS(SMDH_hypothesis_no_comp1, file = here::here("model/SMDH_hypothesis_no_comp1.rds"))

#----------------------------------------------------------------------------
## Now re-running all the 5 meta-regressions once again
#----------------------------------------------------------------------------

# This analysis only includes dataset that has a proxy relating to comaparision type
dataset_lnRR_design_no_comp1 <- dataset_lnRR_no_comp1 %>% filter(!is.na(comparision_type))

VCV_lnRR_design_no_comp1 <- vcalc(vi = lnRR_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_lnRR_design_no_comp1,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )
lnRR_design_no_comp1 <-
  metafor::rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
                    V = VCV_lnRR_design_no_comp1 ,# specify lnRR's sampling variance;
                    mods = ~ - 1 + comparision_type,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                  test = "t",
                  control = list(optimizer="optim"),
                  data = dataset_lnRR_design_no_comp1 )

# Saving the model
saveRDS(lnRR_design_no_comp1, file = here::here("model/lnRR_design_no_comp1.rds"))

# This analysis only includes dataset that has a proxy relating to comaparision type
dataset_SMDH_design_no_comp1 <- dataset_SMDH_no_comp1 %>% filter(!is.na(comparision_type))

VCV_SMDH_design_no_comp1 <- vcalc(vi = SMDH_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_lnRR_design_no_comp1,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )
lnRR_design_no_comp1 <-
  metafor::rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
                    V = VCV_lnRR_design_no_comp1 ,# specify lnRR's sampling variance;
                    mods = ~ - 1 + comparision_type,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                  test = "t",
                  control = list(optimizer="optim"),
                  data = dataset_lnRR_design_no_comp1 )

# Saving the model
saveRDS(lnRR_design_no_comp1, file = here::here("model/lnRR_design_no_comp1.rds"))
```

@@

We have pre-registered the Risk Of Bias analysis, the three we will explore:

1.  **Blinding**

    ```{r}
    unique(dataset_lnRR$blinding)

    lnRR_blinding <-
      metafor::rma.mv(yi = lnRR_sign,
                        V = VCV_lnRR,
                        mods = ~ - 1 + blinding,
                        random = list(~ 1 | paper_ID,
                                      ~ 1 | Observation_ID,
                                      ~ 1 | group_ID_coded,
                                      ~ 1 | population_ID,
                                      ~ 1 | bird_species),
                        method = "REML",
                      test = "t",
                      control = list(optimizer="optim"),
                      data = dataset_lnRR)

    # Saving the model
    saveRDS(lnRR_blinding, file = here::here("model/lnRR_blinding.rds"))

    # Saving the dataset for this model
    write.csv(dataset_lnRR,here::here("data/04_data_analysis/lnRR_blinding.csv"))

    SMDH_blinding <-metafor::rma.mv(yi = SMDH_sign, 
                        V = VCV_SMDH,
                        mods = ~ - 1 + blinding,
                        random = list(~ 1 | paper_ID,
                                      ~ 1 | Observation_ID,
                                      ~ 1 | group_ID_coded,
                                      ~ 1 | population_ID,
                                      ~ 1 | bird_species),
                        method = "REML",
                      test = "t",
                      control = list(optimizer="optim"),
                      data = dataset_SMDH)

    # Saving the model
    saveRDS(SMDH_blinding, file = here::here("model/SMDH_blinding.rds"))

    # Saving the dataset for this model
    write.csv(dataset_SMDH,here::here("data/04_data_analysis/SMDH_blinding.csv"))
    ```

2.  Random Assignment

    unique(dataset_lnRR$random_assignment)

    lnRR_random_assignment <-
      metafor::rma.mv(yi = lnRR_sign,
                        V = VCV_lnRR,
                        mods = ~ - 1 + random_assignment,
                        random = list(~ 1 | paper_ID,
                                      ~ 1 | Observation_ID,
                                      ~ 1 | group_ID_coded,
                                      ~ 1 | population_ID,
                                      ~ 1 | bird_species),
                        method = "REML",
                      test = "t",
                      control = list(optimizer="optim"),
                      data = dataset_lnRR)

    # Saving the model
    saveRDS(lnRR_random_assignment, file = here::here("model/lnRR_random_assignment.rds"))

    # Saving the dataset for this model
    write.csv(dataset_lnRR,here::here("data/04_data_analysis/lnRR_random_assignment.csv"))

    SMDH_random_assignment <-metafor::rma.mv(yi = SMDH_sign, 
                        V = VCV_SMDH,
                        mods = ~ - 1 + random_assignment,
                        random = list(~ 1 | paper_ID,
                                      ~ 1 | Observation_ID,
                                      ~ 1 | group_ID_coded,
                                      ~ 1 | population_ID,
                                      ~ 1 | bird_species),
                        method = "REML",
                      test = "t",
                      control = list(optimizer="optim"),
                      data = dataset_SMDH)

    # Saving the model
    saveRDS(SMDH_random_assignment, file = here::here("model/SMDH_random_assignment.rds"))

    # Saving the dataset for this model
    write.csv(dataset_SMDH,here::here("data/04_data_analysis/SMDH_random_assignment.csv"))
    ```

3.  Missing Vs. Not Missing

    unique(dataset_lnRR$missing_data)

    lnRR_missing_data <-
      metafor::rma.mv(yi = lnRR_sign,
                        V = VCV_lnRR,
                        mods = ~ - 1 + missing_data,
                        random = list(~ 1 | paper_ID,
                                      ~ 1 | Observation_ID,
                                      ~ 1 | group_ID_coded,
                                      ~ 1 | population_ID,
                                      ~ 1 | bird_species),
                        method = "REML",
                      test = "t",
                      control = list(optimizer="optim"),
                      data = dataset_lnRR)

    # Saving the model
    saveRDS(lnRR_missing_data, file = here::here("model/lnRR_missing_data.rds"))

    # Saving the dataset for this model
    write.csv(dataset_lnRR,here::here("data/04_data_analysis/lnRR_missing_data.csv"))

    SMDH_missing_data <-metafor::rma.mv(yi = SMDH_sign, 
                        V = VCV_SMDH,
                        mods = ~ - 1 + missing_data,
                        random = list(~ 1 | paper_ID,
                                      ~ 1 | Observation_ID,
                                      ~ 1 | group_ID_coded,
                                      ~ 1 | population_ID,
                                      ~ 1 | bird_species),
                        method = "REML",
                      test = "t",
                      control = list(optimizer="optim"),
                      data = dataset_SMDH)

    # Saving the model
    saveRDS(SMDH_missing_data, file = here::here("model/SMDH_missing_data.rds"))

    # Saving the dataset for this model
    write.csv(dataset_SMDH,here::here("data/04_data_analysis/SMDH_missing_data.csv"))
    ```

@@

## --- Now for SMDH -----

dataset_SMDH_published<-dataset_SMDH_published%>%
  mutate(year_MeanCenter=scale(as.numeric(year_publication),scale=F)[,1])

@@

```{r}
# ## For lnRR complete dataset
# 
# ## I am changing the year of unpublished studies to 2025 which is to now, assuming that if they were to be published now, what would be the effect. I will keep in mind that in this model if I do find an evidence of publication bias over years, I should keep in mind it is non-independent of the unpublished studies
# 
dataset_lnRR_all_in<-dataset_lnRR%>%mutate(year_publication = case_when(is.na(year_publication) ~ 2025, TRUE ~ year_publication))%>%
  mutate(year_MeanCenter=scale(as.numeric(year_publication),scale=F)[,1])%>%
  mutate(effective_n_inv=1/((4*effective_n_experiment*effective_n_control)/(effective_n_experiment+effective_n_control)))%>%
  mutate(sqrt_effective_n_inv=sqrt(effective_n_inv))%>%
  mutate(comparision_type= case_when(is.na(comparision_type) ~ as.factor(4), TRUE ~ comparision_type))

VCV_lnRR_all_in<- vcalc(vi = lnRR_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_lnRR_all_in,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )

lnRR_all_in <- rma.mv(yi = lnRR_sign, # specify lnRR as the effect size measure;
                    V = VCV_lnRR_all_in, # specify lnRR's sampling variance-covariance matrix;
                    mods = ~ time_of_gnm_addition +
                      year_MeanCenter +
                      sqrt_effective_n_inv +
                      comparision_type +
                      bird_species +
                      trait_type,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                    control = list(optimizer="optim"),
                    test = "t",
                    data = dataset_lnRR_all_in)

# saving the model
saveRDS(lnRR_all_in, file = here::here("model/lnRR_all_in.rds"))

## For SMDH complete published dataset

dataset_SMDH_all_in<-dataset_SMDH%>%mutate(year_publication = case_when(is.na(year_publication) ~ 2025, TRUE ~ year_publication))%>%
  mutate(year_MeanCenter=scale(as.numeric(year_publication),scale=F)[,1])%>%
  mutate(effective_n_inv=1/((4*effective_n_experiment*effective_n_control)/(effective_n_experiment+effective_n_control)))%>%
  mutate(sqrt_effective_n_inv=sqrt(effective_n_inv))%>%
  mutate(comparision_type= case_when(is.na(comparision_type) ~ as.factor(4), TRUE ~ comparision_type))

VCV_SMDH_all_in<- vcalc(vi = SMDH_variance,
                  cluster = paper_ID,
                  obs = Observation_ID,
                  data = dataset_SMDH_all_in,
                  rho = 0.5 # assuming that the effect sizes within the same study are correlated with rho = 0.5
                  )

SMDH_all_in<- rma.mv(yi = SMDH_sign, # specify lnRR as the effect size measure;
                    V = VCV_SMDH_all_in, # specify lnRR's sampling variance-covariance matrix;
                    mods = ~ time_of_gnm_addition +
                      year_MeanCenter +
                      sqrt_effective_n_inv +
                      as.factor(comparision_type) +
                      bird_species +
                      trait_type,
                    random = list(~ 1 | paper_ID,
                                  ~ 1 | Observation_ID,
                                  ~ 1 | group_ID_coded,
                                  ~ 1 | population_ID,
                                  ~ 1 | bird_species),
                    method = "REML",
                    control = list(optimizer="optim"),
                    test = "t",
                    data = dataset_SMDH_all_in)

# saving the model
saveRDS(SMDH_all_in, file = here::here("model/SMDH_all_in.rds"))
```

@@

Including post-hoc Wald tests

For this I have to run all the meta-regressions again without removing the intercepts
